Pomace: a Grappa for Non-Volatile Memory \vspace{-.2in}

### 2012-12-11 doc/nvm2013/paper.tex
holt's comments on sec 1 & a bit of whitespace removal

\title{Pomace: a Grappa for Non-Volatile Memory}
\authorinfo{Jacob Nelson$^{\dagger}$,
  Brandon Holt$^{\dagger}$,
  Brandon Myers$^{\dagger}$,
  Preston Briggs$^{\dagger}$,
  Luis Ceze$^{\dagger}$,
  Simon Kahan$^{{\dagger \ddagger}}$,
  Mark Oskin$^{\dagger}$
}{\textdagger University of Washington, \textdaggerdbl Pacific Northwest National Laboratory}{
  \{nelson, bholt, bdmyers, preston, luisceze, skahan, oskin\}@cs.washington.edu}

\maketitle
\begin{abstract}
We revisit Grappa, our software platform that scales in-memory irregular
applications on commodity clusters.
We contemplate options for adapting Grappa to use non-volatile memory technologies.
\end{abstract}

@@

\item[Distributed shared memory.] Our DSM system provides support for
fine-grain access to data anywhere in the system. It supports synchronization
operations on global data, explicit local caching of any memory in the system,
and support for operations on remote data (delegating operations to home node).
By tight integration with the tasking system and the
communication layer, our DSM system offers high aggregate random
access bandwidth to remote data.

@@

@@

Grappa thus facilitates
expression of fine-grained task parallelism at the application
programming level, employing that concurrency as needed to sustain throughput in the presence of long latency operations and to better utilize commodity networks.  Programmers attend only to expressing the parallelism in their program, not to scheduling it.  While similar latency
tolerance and aggregation techniques can be manually coded into some
applications, Grappa relieves the programmer of this burden.  In
addition, Grappa provides an API for reader-writer semantics for the
purpose of manipulating a local copy of any contiguous portion of the
shared address space, enabling programmers to exploit spatial and
temporal locality while maintaining a consistent global view.  With
these features, we have demonstrated performance on commodity systems
for highly irregular applications approaching that of the Cray XMT, a
fully custom system.

@@

\section{Applying non-volatile memory to Grappa}
The natural point of integration for non-volatile memories in Grappa
is to store the global heap in nonvolatile memory, since the heap
usually contains data that is infrequently modified. All accesses to
this shared heap in Grappa programs are done through an API. If the
access is to local memory, Grappa issues a prefetch for the address
and context-switches to other work while the prefetch is executing. If
the access is to memory stored on another node, Grappa issues a
message to that node requesting the data and context-switches to other
work until the reply arrives. Non-volatile memory would be accessed in
the same way: if a request is to memory stored in non-volatile memory,
Grappa would initiate the request and context-switch to other work
until the request had completed.

@@

\paragraph{SSDs have a low request rate.} The fastest (cite) can
perform on the order of 1 million random reads per second, and many
are limited to less than 100K/s. In contrast, cores may be idle in
Grappa if each node cannot make 100 million random reads per second of
the global heap. Even the individual flash chips are slow: 50K/s (cite
micron).

\paragraph{SSDs have high overhead.} Commodity SSDs are usually treated
as normal disks---accesses must occur through the kernel, incurring a
high context switch overhead. Grappa depends on user-mode access to
the network layer. This capability has been explored in NVM research
(cite Moneta).

@@

For simplicity, we omit discussion of write and erase operations. We
assume a non-volatile Grappa would store frequently-written temporary
data in DRAM and would use its standard latency-tolerance techniques
to cover the cost of infrequent NVM writes.

@@

There are two problems with this design. First, while each node
requires fewer chips, the overall cluster performance goal, and thus
chip requirement, is the same. Second, low single-threaded performance is detrimental to performance during periods of diminished application concurrency.

@@

While similar techniques have provided benefit in prior work in
restricted situations (cite), they depend on a sufficient request
density to each block to get a benefit. As datasets scale up, the
request density may decrease and the benefit of this technique may
decrease.

@@

\paragraph{PCM as Main Memory}
Much recent research has explored the use of PCM as a main memory
replacement. Grappa would easily be able to take advantage of such a
development. When a Grappa program references a low-locality program,
a prefetch request is issued for that data a few cycles before the
thread is context-switched in\cite{Nelson:hotpar2011}. A PCM-based
Grappa would do the same, but would issue the prefetches earlier to
cover the longer latency of PCM.

@@

Grappa could also support a PCM array in a separate address space (for
instance, on the PCIe bus). PCM chips available today (cite) support a
64-bit read in 200 ns; this means that 20 chips acting independently
would be sufficient to reach our target request rate.

@@

An ideal array would support limited atomic operations on 64-bit
words, including increment and compare-and-swap.

>>>>

\title{Pomace: a Grappa for Non-Volatile Memory \vspace{-.2in}}
\authorinfo{Jacob Nelson$^{\dagger}$,
  Brandon Holt$^{\dagger}$,
  Brandon Myers$^{\dagger}$,
  Preston Briggs$^{\dagger}$,
  Luis Ceze$^{\dagger}$,
  Simon Kahan$^{{\dagger \ddagger}}$,
  Mark Oskin$^{\dagger}$
}{\textdagger University of Washington, \textdaggerdbl Pacific Northwest National Laboratory}{
  \{nelson, bholt, bdmyers, preston, luisceze, skahan, oskin\}@cs.washington.edu \vspace{-1in}
}

\maketitle
\begin{abstract}
We revisit Grappa, our software platform that scales in-memory irregular
applications on commodity clusters.
We contemplate options for adapting Grappa to use non-volatile memory technologies.
\end{abstract}
\vspace{-.1in}

@@

@@

\item[Distributed shared memory.] Our DSM system provides support for
fine-grain access to data anywhere in the system. It supports synchronization
operations on global data, explicit local caching of any memory in the system,
and support for operations on remote data (delegating operations to owner node).
By tight integration with the tasking system and the
communication layer, our DSM system offers high aggregate random
access bandwidth to remote data.

@@

Grappa thus facilitates
expression of fine-grained task parallelism at the application
programming level, employing that concurrency as needed to sustain throughput in the presence of long latency operations and to better utilize commodity networks.  Programmers attend only to expressing the parallelism in their program, not to scheduling it.  While similar latency
tolerance and aggregation techniques can be manually coded into some
applications, Grappa relieves the programmer of this burden.  In
addition, Grappa provides an API for reader-writer lock semantics for the
purpose of manipulating a local copy of any contiguous portion of the
shared address space, enabling programmers to exploit spatial and
temporal locality while maintaining a consistent global view.  With
these features, we have demonstrated performance on commodity systems
for highly irregular applications approaching that of the Cray XMT, a
fully custom system.

@@

\section{Applying non-volatile memory to Grappa}
The natural point of integration for non-volatile memories in Grappa
is to store the global heap in nonvolatile memory, since the heap
usually contains data that is infrequently modified. All accesses to
this shared heap in Grappa programs are done through an API. If the
access is to local memory, Grappa issues a prefetch for the address
and context-switches to other work while the prefetch is executing~\cite{Nelson:hotpar2011}. If
the access is to memory stored on another node, Grappa issues a
message to that node requesting the data and context-switches to other
work until the reply arrives. Non-volatile memory would be accessed in
the same way: if a request is to memory stored in non-volatile memory,
Grappa would initiate the request and context-switch to other work
until the request had completed.

@@

\paragraph{SSDs have a low request rate.} The fastest~\cite{fusionio} can
perform on the order of 1 million random reads per second, and many
are limited to less than 100K/s. In contrast, cores may be idle in
Grappa if each node cannot make 100 million random reads per second of
the global heap. Even the individual flash chips are slow, on the order of 40K/s~\cite{micronFlash}.

\paragraph{SSDs have high overhead.} Commodity SSDs are usually treated
as normal disks---accesses must occur through the kernel, incurring a
high context switch overhead. Grappa depends on user-mode access to
the network layer. This capability has been explored in recent research~\cite{caulfield:2012}.

@@

For simplicity, we omit discussion of write and erase operations. We
assume a non-volatile Grappa would store frequently-written temporary
data in DRAM and would use its standard latency-tolerance techniques
to cover the cost of infrequent writes.

@@

There are two problems with this design for our purpose. First, while each node
requires fewer chips, the overall cluster performance goal, and thus
chip requirement, is the same. Second, low single-threaded performance is detrimental to performance during periods of diminished application concurrency.

@@

While similar techniques have provided benefit in prior work in
restricted situations~\cite{magt:2010,graphchi:osdi12}, they depend on a sufficient request
density to each block to get a benefit. As datasets scale up, the
request density may decrease and the benefit of this technique may
decrease.

@@

\paragraph{PCM as Main Memory}
Much recent research has explored the use of PCM as a main memory
replacement. Grappa would easily be able to take advantage of such a
development. When a Grappa program references a low-locality program,
a prefetch request is issued for that data a few cycles before the
thread is context-switched in. A PCM-based
Grappa would do the same, but would issue the prefetches earlier to
cover the longer latency of PCM.

@@

Grappa could also support a PCM array in a separate address space (for
instance, on the PCIe bus). PCM chips available today~\cite{micronPCM} support a
64-bit read in 200 ns; this means that 20 chips acting independently
would be sufficient to reach our target request rate.

@@

%An ideal array would support limited atomic operations on 64-bit
%words, including increment and compare-and-swap.
